\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-29 paper.qmd
done with background

The solution is to use the tour to examine how the NLDR is warping the space. This approach follows what @article08 describes as *model-in-the-data-space*. The fitted model should be overlaid on the

Linked brushing as done by @article21

- Model-in-the-data-space: how can we represent the model, eg plane for PCA, grid of values for classification boundaries.

>>>>

The solution is to use the tour to examine how the NLDR is warping the space. This approach follows what @article08 describes as *model-in-the-data-space*. The fitted model should be overlaid on the data, to examine the fit relative the spread of the observations. While this is straightforward, and commonly done when data is 2D, it is also possible in $p-D$, for many models, when a tour is used.

@article08 provides several examples of models overlaid on the data in $p-D$. In hierarchical clustering, a representation of the dendrogrom using points and lines can be constructed by augmenting the data with points marking merging of clusters. Showing the movie of linear projections reveals shows how the algorithm sequentially fitted the cluster model to the data. For linear discriminant analysis or model-based clustering the model can be indicated by $(p-1)-D$ ellipses. It is possible to see whether the elliptical shapes appropriately matches the variance of the relevant clusters, and to compare and contrast different fits. For PCA, one can display the $k-D$ plane of the reduced dimension using wireframes of transformed cubes. Using a wireframe is the approach we take here, to represent the NLDR model in $p-D$.

<!-- Linked brushing as done by @article21

- Model-in-the-data-space: how can we represent the model, eg plane for PCA, grid of values for classification boundaries, ellipses for LDA and mclust, nets for SOM.-->

### 2024-03-29 paper.tex
done with background

The solution is to use the tour to examine how the NLDR is warping the
space. This approach follows what \citet{article08} describes as
\emph{model-in-the-data-space}.

\begin{itemize}
\tightlist
\item
  Model-in-the-data-space: how can we represent the model, eg plane for
  PCA, grid of values for classification boundaries.
\end{itemize}

>>>>

The solution is to use the tour to examine how the NLDR is warping the
space. This approach follows what \citet{article08} describes as
\emph{model-in-the-data-space}. The fitted model should be overlaid on
the data, to examine the fit relative the spread of the observations.
While this is straightforward, and commonly done when data is 2D, it is
also possible in \(p-D\), for many models, when a tour is used.

\citet{article08} provides several examples of models overlaid on the
data in \(p-D\). In hierarchical clustering, a representation of the
dendrogrom using points and lines can be constructed by augmenting the
data with points marking merging of clusters. Showing the movie of
linear projections reveals shows how the algorithm sequentially fitted
the cluster model to the data. For linear discriminant analysis or
model-based clustering the model can be indicated by \((p-1)-D\)
ellipses. It is possible to see whether the elliptical shapes
appropriately matches the variance of the relevant clusters, and to
compare and contrast different fits. For PCA, one can display the
\(k-D\) plane of the reduced dimension using wireframes of transformed
cubes. Using a wireframe is the approach we take here, to represent the
NLDR model in \(p-D\).
